Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-11-08 paper/ms.tex
some words in the intro

Stars are variable, for better or worse:  their variability provides information about the properties
of stars, but also results in noise that must be accounted for in detecting and characterizing their
planetary systems.  When analyzing astronomical times series of stars, then, there are generally two
goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to better understand stellar
properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
flux variability and spectral variations.

A promising technique for addressing both of these goals is the Gaussian Process formalism
\citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
correlations is completely specified by a correlation matrix.  In its simplest form, the
correlation matrix is modeled with a time-independent autocorrelation function (although
the time-independence can be relaxed), including a white-noise component that may vary
with each exposure, and the autocorrelation function is described by a functional form that may be
simply parameterized by a relatively small number of parameters, sometimes referred to as
`hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
describe the autocorrelation function, thus inferring properties of the stellar variabilty
which may be used to characterize properties of the star, such as the rotation period,
the asteroseismic frequencies, or granulation noise known as `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
The goals of planet detection include measuring the Doppler shift of the star, measuring
the decrement of flux as a planet transits a star, and/or measuring the times of transit to
look for dynamical interactions between planets.  To properly account for the impact of
stellar variability on the planet parameters requires treating correlations in the data, and
to do so also requires evaluating the correlation matrix.

The size of the correlation matrix, $N_{elements}$, scales as the square of the length of the time-series,
$N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become
prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
= N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
using long time series datsets, such as the {\it Kepler} dataset $N_{time} \approx 10^{5-6}$,
{\it Spitzer} time series with $N_{time} \approx 10^5$, and Solar time series with
$N_{time} \approx 10^7$.  So a faster means of solving for the likelihood is required in these
cases.

One promising solution is to use approximate techiques based upon the fact that correlation
matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the
solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
solver is complicated and can be challenging to use with general choices of kernel.  Another
promising approximate technique is to use a uniform resampling of the dataset, and then
rely on the structure of Toeplitz matrices, which require uniformly space time series,
to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
this approach has the drawback of being approximate.  A third approach is to use a kernel
based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
of the Lorentzians are required to have a specific relation.  In addition, these
techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}

The approach we explore in this paper is based upon a method developed by Press \&
Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
They also indicated that a combination of two exponential kernels would have a simliar scaling,
but they did not demonstrate how to extend this to more than two kernel components.

A significant advance was made in generalizing the Press-Rybicki method to an
arbitrary number of exponential kernels by utilizing the fact that the elements of
such matrices can be written in terms of the products of components of two vectors,
which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
Rybicki-Press' (\texttt{GenRP}) approach
has many advantages:  the computation time and storage both scale as $O(N_{time})$,
the coefficients of the kernels need not be stationary, the computation of the
likelihood is exact, and the algorithm is simple to implement, and thus is robust.
Unlike the \texttt{CARMA} model, there is no restriction on the relation between
the coefficients (except that the kernel needs to be positive definite).  Unlike
\texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback
of this approach is that the sum of exponential kernels with real coefficients can
only represent a limited range of power spectra of auto-correlated time series:
those which can be expressed the sum of Lorentzians that peak at zero frequency.

However, this drawback turns out to not be significant:  the exponential kernel
can also have a complex exponential coefficients (plus their complex-conjugates), and
thus approximate an almost arbitrary autocorrelation function with enough terms, most
importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
components is a {\it very} good approximation of stellar variability spectra:  the
activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
Lorentzians \citep{1985ESASP.235..199H}, the asteroseismic oscillations may be expressed as non-zero frequency
Lorentzians \citep{1990ApJ...364..699A,Gruberbauer2009}, and stellar rotation may be
parameterized this way as well.  In this paper
we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when
combined with the complex conjugate leads to damped harmonic auto-correlation functions
which are flexible enough to describe a wide range of variability, and may still
be solved in $O(N_{time}$ operations.

We show that the extended matrix that embeds this multi-component kernel can be written
in real notation, which speeds up the evaluation of the likelihood function,
and allows optimization software to use automatic differential in computing
derivatives of the likelihood function.

The plan of the paper is...

>>>>

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars (CITE), active galactic nuclei (CITE),
and X-ray binaries (CITE).
They have also been used as models for the cosmic microwave background (CITE),
correlated instrumental noise (CITE), spectroscopic calibration (CITE) and
residuals caused by model inconsistencies (CITE + better words).
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the third power of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time-domain surveys~--~with $\sim10^{4-9}$ targets with $\sim10^{3-5}$
observations each~---~this cost is prohibitive.

In this paper, we present a class of GP models that enable likelihood
calculations that scale linearly with the number of data points
$\mathcal{O}(N)$ for one dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must be written as a mixture of Lorentzian functions in
Fourier space.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic.
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, robust, simple,
and fast.

In the following pages, we will motivate the general problem of GP regression,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released efficient and well-tested
implementations of this method written in \project{C++}, \project{Python}, and
\project{Julia}.
These implementations are available online at \project{GitHub}
\url{https://github.com/dfm/GenRP} and \project{Zenodo} \dfmtodo{add zenodo
archive}.

% DFM: I'm going to redistribute the following in other sections.

% Stars are variable, for better or worse:  their variability provides information about the properties
% of stars, but also results in noise that must be accounted for in detecting and characterizing their
% planetary systems.  When analyzing astronomical times series of stars, then, there are generally two
% goals:  1). to analyze the frequency spectrum and amplitude of stellar variability to better understand stellar
% properties and evolution, and/or 2). to account for correlated noise when modeling the stellar
% flux variability and spectral variations.

% A promising technique for addressing both of these goals is the Gaussian Process formalism
% \citep{Rasmussen2006,Gibson2012}.  The basis of this method is to assume that stellar variability
% is random, Gaussian noise, but that the random variability is correlated in time, and the nature of the
% correlations is completely specified by a correlation matrix.  In its simplest form, the
% correlation matrix is modeled with a time-independent autocorrelation function (although
% the time-independence can be relaxed), including a white-noise component that may vary
% with each exposure, and the autocorrelation function is described by a functional form that may be
% simply parameterized by a relatively small number of parameters, sometimes referred to as
% `hyper-parameters.'  The goal of spectral analysis is to measure the parameters which
% describe the autocorrelation function, thus inferring properties of the stellar variabilty
% which may be used to characterize properties of the star, such as the rotation period,
% the asteroseismic frequencies, or granulation noise known as `flicker' \citep{Aerts2010,Noyes1984,Bastien2013}.
% The goals of planet detection include measuring the Doppler shift of the star, measuring
% the decrement of flux as a planet transits a star, and/or measuring the times of transit to
% look for dynamical interactions between planets.  To properly account for the impact of
% stellar variability on the planet parameters requires treating correlations in the data, and
% to do so also requires evaluating the correlation matrix.

% The size of the correlation matrix, $N_{elements}$, scales as the square of the length of the time-series,
% $N_{element}=N_{time}^2$, so as time series grow in size, say $N_{time} =10^{5-6}$, the storage may become
% prohibitive, $N_{element} = 10^{10-12}$.  In addition, this matrix must be inverted and have
% its determinant evaluated to compute the likelihood function, which takes of order $N_{operations}
% = N_{time}^3 = 10^{15-18}$.  This becomes an impossible computational problem, which prohibits
% using long time series datsets, such as the {\it Kepler} dataset $N_{time} \approx 10^{5-6}$,
% {\it Spitzer} time series with $N_{time} \approx 10^5$, and Solar time series with
% $N_{time} \approx 10^7$.  So a faster means of solving for the likelihood is required in these
% cases.

% One promising solution is to use approximate techiques based upon the fact that correlation
% matrices display a great deal of symmetry.  This `hierarchical off-diagonal low-rank' (\texttt{HODLR})
% algorithm can make the storage and operations both scale in proportion to $N_{time}$, although the
% solver is only approximate \citep{Ambikasaran2013,Ambikasaran2016}.  In addition, the
% solver is complicated and can be challenging to use with general choices of kernel.  Another
% promising approximate technique is to use a uniform resampling of the dataset, and then
% rely on the structure of Toeplitz matrices, which require uniformly space time series,
% to obtain approximate solutions of the likelihood function \citep{WilsonNickisch2015},
% the `Kernel Interpolation for Scalable Structured Gaussian Processes' (\texttt{KISS-GP});
% this approach has the drawback of being approximate.  A third approach is to use a kernel
% based upon stochastic differential equations:  the `continuous auto-regressive moving-average'
% model \citep[aka \texttt{CARMA};][]{Kelly2014}.  This approach can model a stationary power-spectrum that
% is the combination of a number of Lorentzians, but has the disadvantage that the amplitudes
% of the Lorentzians are required to have a specific relation.  In addition, these
% techniques require a stationary kernel. {\color{red} Is this true for GEORGE?}

% The approach we explore in this paper is based upon a method developed by Press \&
% Rybicki twenty years ago \citep{1995PhRvL..74.1060R}.  They observed that a correlation
% matrix consisting of a single exponential kernel has an inverse that is tri-diagonal.
% The coefficients and solution of a tri-diagonal matrix scale in proportion to $O(N_{time})$.
% They also indicated that a combination of two exponential kernels would have a simliar scaling,
% but they did not demonstrate how to extend this to more than two kernel components.

% A significant advance was made in generalizing the Press-Rybicki method to an
% arbitrary number of exponential kernels by utilizing the fact that the elements of
% such matrices can be written in terms of the products of components of two vectors,
% which are so-called semi-separable matrices \citep{Ambikasaran2015}.  This `Generalized
% Rybicki-Press' (\texttt{GenRP}) approach
% has many advantages:  the computation time and storage both scale as $O(N_{time})$,
% the coefficients of the kernels need not be stationary, the computation of the
% likelihood is exact, and the algorithm is simple to implement, and thus is robust.
% Unlike the \texttt{CARMA} model, there is no restriction on the relation between
% the coefficients (except that the kernel needs to be positive definite).  Unlike
% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback
% of this approach is that the sum of exponential kernels with real coefficients can
% only represent a limited range of power spectra of auto-correlated time series:
% those which can be expressed the sum of Lorentzians that peak at zero frequency.

% However, this drawback turns out to not be significant:  the exponential kernel
% can also have a complex exponential coefficients (plus their complex-conjugates), and
% thus approximate an almost arbitrary autocorrelation function with enough terms, most
% importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
% components is a {\it very} good approximation of stellar variability spectra:  the
% activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
% Lorentzians \citep{1985ESASP.235..199H}, the asteroseismic oscillations may be expressed as non-zero frequency
% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer2009}, and stellar rotation may be
% parameterized this way as well.  In this paper
% we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when
% combined with the complex conjugate leads to damped harmonic auto-correlation functions
% which are flexible enough to describe a wide range of variability, and may still
% be solved in $O(N_{time}$ operations.

% We show that the extended matrix that embeds this multi-component kernel can be written
% in real notation, which speeds up the evaluation of the likelihood function,
% and allows optimization software to use automatic differential in computing
% derivatives of the likelihood function.

% The plan of the paper is...
